---
title: "getting started"
---

```{r setup-library}
library(handsontable)
```


## basic

```{r basic-table}
handsontable(data = iris[1:10, ], adaptiveHeight = TRUE)
```

## read-only

to make the table read-only, set `readOnly = TRUE`. this will prevent any editing of the table.

```{r read-only-table}
handsontable(
  data = iris[1:5, ],
  adaptiveHeight = TRUE,
  readOnly = TRUE
)
```

## headers

- colHeaders

  by default, `colHeaders = TRUE` and uses the column names of the data frame.

  ```{r colheaders-true}
  handsontable(
    data = iris[1:5, ],
    adaptiveHeight = TRUE,
    colHeaders = TRUE
  )
  ```

  setting `colHeaders = FALSE` will remove the column headers.

  ```{r colheaders-false}
  handsontable(
    data = iris[1:5, ],
    adaptiveHeight = TRUE,
    colHeaders = FALSE
  )
  ```

  to set custom column headers, pass a character vector to `colHeaders`.

  ```{r colheaders-custom}
  handsontable(
    data = iris[1:5, ],
    adaptiveHeight = TRUE,
    colHeaders = c(
      "Sepal Length",
      "Sepal Width",
      "Petal Length",
      "Petal Width",
      "Species"
    )
  )
  ```

- rowHeaders

  by default, `rowHeaders = TRUE` and uses the row names of the data frame.

  ```{r rowheaders-true}
  handsontable(
    data = mtcars[1:5, ],
    adaptiveHeight = TRUE,
    rowHeaders = TRUE,
    rowHeaderWidth = 200
  )
  ```

  setting `rowHeaders = FALSE` will remove the row headers.

  ```{r rowheaders-false}
  handsontable(
    data = mtcars[1:5, ],
    adaptiveHeight = TRUE,
    rowHeaders = FALSE
  )
  ```

  to set custom row headers, pass a character vector to `rowHeaders`.

  ```{r rowheaders-custom}
  handsontable(
    data = mtcars[1:5, ],
    adaptiveHeight = TRUE,
    rowHeaders = c("Row 1", "Row 2", "Row 3", "Row 4", "Row 5")
  )
  ```


## dimensions

- width & height

  ```{r dimensions-width-height}
  handsontable(
    data = mtcars,
    width = 500,
    height = 300,
    rowHeaderWidth = 200
  )
  ```

- setting `adaptiveHeight = TRUE` will make the table height adapt to the 
content of the cells, so that no vertical scrollbar is shown.
  ```{r dimensions-adaptive-height}
  handsontable(
    data = mtcars,
    adaptiveHeight = TRUE,
    rowHeaderWidth = 200
  )
  ```

## column widths

to set the width of individual columns, pass a numeric vector to `colWidths`.

```{r column-widths}
handsontable(
  data = iris[1:5, ],
  adaptiveHeight = TRUE,
  colWidths = c(100, 150, 200, 250, 300)
)
```

## column stretching

- `stretchH = "all"` will stretch all columns to fit the width of the table.
  ```{r stretch-all}
  handsontable(
    data = iris[1:5, 1:3],
    adaptiveHeight = TRUE,
    stretchH = "all"
  )
  ```
- `stretchH = "last"` will stretch the last column to fit the width of the table.
  ```{r stretch-last}
  handsontable(
    data = iris[1:5, 1:3],
    adaptiveHeight = TRUE,
    stretchH = "last"
  )
  ```
- `stretchH = "none"` will not stretch any columns.
  ```{r stretch-none}
  handsontable(
    data = iris[1:5, 1:3],
    adaptiveHeight = TRUE,
    stretchH = "none"
  )
  ```


## fixed columns & rows

- set `fixedColumnsLeft` to fix a number of columns on the left side of the table.
  ```{r fixed-columns-left}
  handsontable(
    data = mtcars[1:10, ],
    adaptiveHeight = TRUE,
    fixedColumnsLeft = 2,
    rowHeaderWidth = 200,
    width = 500
  )
  ```
- set `fixedRowsTop` to fix a number of rows on the top of the table.
  ```{r fixed-rows-top}
  handsontable(
    data = mtcars,
    fixedRowsTop = 2,
    rowHeaderWidth = 200
  )
  ```
- set `fixedRowsBottom` to fix a number of rows on the bottom of the table.
  ```{r fixed-rows-bottom}
  handsontable(
    data = mtcars,
    fixedRowsBottom = 2,
    rowHeaderWidth = 200
  )
  ```

## manual column & row move

- setting `manualColumnMove = TRUE` will allow users to move columns by dragging the column headers.
  ```{r manual-column-move}
  handsontable(
    data = iris[1:10, ],
    adaptiveHeight = TRUE,
    manualColumnMove = TRUE
  )
  ```
- setting `manualRowMove = TRUE` will allow users to move rows by dragging the row headers.
  ```{r manual-row-move}
  handsontable(
    data = mtcars,
    adaptiveHeight = TRUE,
    manualRowMove = TRUE,
    rowHeaderWidth = 200
  )
  ```
