---
title: "getting started"
---

```{r}
library(handsontable)
```


## basic

```{r}
handsontable(data = mtcars)
```

## read-only

to make the table read-only, set `readOnly = TRUE`. this will prevent any editing of the table.

```{r}
handsontable(data = mtcars, readOnly = TRUE)
```

## headers

- colHeaders

  by default, `colHeaders = TRUE` and uses the column names of the data frame.

  ```{r}
  handsontable(data = mtcars, colHeaders = TRUE)
  ```

  setting `colHeaders = FALSE` will remove the column headers.

  ```{r}
  handsontable(data = mtcars, colHeaders = FALSE)
  ```

  to set custom column headers, pass a character vector to `colHeaders`.

  ```{r}
  handsontable(
    data = mtcars,
    colHeaders = c(
      "Miles per Gallon",
      "Cylinders",
      "Displacement",
      "Horsepower",
      "Rear Axle Ratio",
      "Weight (1000 lbs)",
      "1/4 Mile Time",
      "Engine",
      "Transmission",
      "Number of forward gears",
      "Number of carburetors"
    )
  )
  ```

- rowHeaders

  by default, `rowHeaders = TRUE` and uses the row names of the data frame.

  ```{r}
  handsontable(data = mtcars, rowHeaders = TRUE)
  ```

  setting `rowHeaders = FALSE` will remove the row headers.

  ```{r}
  handsontable(data = mtcars, rowHeaders = FALSE)
  ```

  to set custom row headers, pass a character vector to `rowHeaders`.

  ```{r}
  handsontable(
    data = mtcars,
    rowHeaders = c("Row 1", "Row 2", "Row 3", "Row 4", "Row 5")
  )
  ```


## dimensions

- width & height

  ```{r}
  handsontable(data = mtcars, width = 300, height = 300)
  ```

- setting `adaptiveHeight = TRUE` will make the table height adapt to the 
content of the cells, so that no vertical scrollbar is shown.
  ```{r}
  handsontable(data = mtcars, adaptiveHeight = TRUE)
  ```

## column widths

to set the width of individual columns, pass a numeric vector to `colWidths`.

```{r}
handsontable(data = mtcars, colWidths = c(100, 150, 200, 250, 300, 350, 400, 450))
```

## column stretching

- `stretchH = "all"` will stretch all columns to fit the width of the table.
  ````{r}
  handsontable(data = mtcars[, 1:3], stretchH = "all")
  ````
- `stretchH = "last"` will stretch the last column to fit the width of the table.
  ```{r}
  handsontable(data = mtcars[, 1:3], stretchH = "last")
  ```
- `stretchH = "none"` will not stretch any columns.
  ```{r}
  handsontable(data = mtcars[, 1:3], stretchH = "none")
  ```


## fixed columns & rows

- set `fixedColumnsLeft` to fix a number of columns on the left side of the table.
  ```{r}
  handsontable(data = mtcars, fixedColumnsLeft = 2, width = 300)
  ```
- set `fixedRowsTop` to fix a number of rows on the top of the table.
  ```{r}
  handsontable(data = mtcars, fixedRowsTop = 2)
  ```
- set `fixedRowsBottom` to fix a number of rows on the bottom of the table.
  ```{r}
  handsontable(data = mtcars, fixedRowsBottom = 2)
  ```

## manual column & row resize

- setting `manualColumnResize = TRUE` will allow users to resize columns by dragging the column borders.
  ```{r}
  handsontable(data = mtcars, manualColumnResize = TRUE)
  ```

- setting `manualRowResize = TRUE` will allow users to resize rows by dragging the row borders.
  ```{r}
  handsontable(data = mtcars, manualRowResize = TRUE)
  ```

## manual column & row move

- setting `manualColumnMove = TRUE` will allow users to move columns by dragging the column headers.
  ```{r}
  handsontable(data = mtcars, manualColumnMove = TRUE)
  ```
- setting `manualRowMove = TRUE` will allow users to move rows by dragging the row headers.
  ```{r}
  handsontable(data = mtcars, manualRowMove = TRUE)
  ```
