---
title: "Configuration Options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Configuration Options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = interactive()
)
```

```{r setup}
library(handsontable)
```

# Configuration Options

This vignette covers the various configuration functions available in the 
handsontable package, allowing you to customize every aspect of your tables.

## General Table Configuration

The `handsontable()` function controls overall table behavior and features.

### Context Menu and Interaction

```{r context-menu}
# Enable context menu and other interactive features
handsontable(
  data = iris[1:8, ],
  filters = TRUE,
  manualColumnSorting = TRUE,
  search = TRUE,
  undo = TRUE
) |>
  hot_context_menu()
```

### Multiple Column Sorting

```{r multi-sort}
# Enable multi-column sorting
handsontable(
  data = mtcars[1:10, 1:6],
  multiColumnSorting = TRUE,
  outsideClickDeselects = TRUE
)
```

### Fill Handle Configuration

```{r fill-handle}
# Configure fill handle behavior
handsontable(
  data = data.frame(
    Numbers = 1:5,
    Letters = letters[1:5]
  ),
  fillHandle = list(
    direction = "vertical",
    autoInsertRow = TRUE
  )
)
```

## Individual Column Configuration with `hot_col()`

Use `hot_col()` to configure specific columns with different types and 
behaviors.

### Text Columns

```{r text-columns}
# Configure a text column
data <- data.frame(
  ID = 1:5,
  Name = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  Notes = rep("", 5)
)

handsontable(data) |>
  hot_col(col = 1, readOnly = TRUE, width = 50) |>
  hot_col(col = 2, type = "text", width = 100) |>
  hot_col(col = 3, type = "text", width = 200)
```

### Numeric Columns

```{r numeric-columns}
# Configure numeric columns with formatting
financial_data <- data.frame(
  Product = c("Widget A", "Widget B", "Widget C"),
  Price = c(19.99, 24.50, 15.75),
  Quantity = c(100, 250, 75)
)

handsontable(financial_data) |>
  hot_col(col = 2, type = "numeric", format = "$0.00") |>
  hot_col(col = 3, type = "numeric", format = "0,0")
```

### Dropdown Columns

```{r dropdown-columns}
# Create dropdown columns
employee_data <- data.frame(
  Name = c("John", "Jane", "Bob"),
  Department = c("IT", "HR", "Finance"),
  Level = c("Senior", "Junior", "Manager"),
  stringsAsFactors = FALSE
)

handsontable(employee_data) |>
  hot_col(
    col = 2,
    type = "dropdown",
    source = c("IT", "HR", "Finance", "Marketing", "Sales"),
    strict = TRUE
  ) |>
  hot_col(
    col = 3,
    type = "dropdown",
    source = c("Junior", "Senior", "Manager", "Director"),
    strict = TRUE
  )
```

### Autocomplete Columns

```{r autocomplete-columns}
# Create autocomplete columns (like dropdown but with filtering)
country_data <- data.frame(
  City = c("New York", "London", "Tokyo"),
  Country = c("USA", "UK", "Japan"),
  Region = c("North America", "Europe", "Asia"),
  stringsAsFactors = FALSE
)

countries <- c("USA", "UK", "Japan", "Germany", "France", "Canada", "Australia", "Brazil")
regions <- c("North America", "Europe", "Asia", "Oceania", "South America")

handsontable(country_data) |>
  hot_col(
    col = 2,
    type = "autocomplete",
    source = countries,
    strict = FALSE, # Allow new values
    filter = TRUE, # Enable filtering as you type
    trimDropdown = TRUE # Fit dropdown to content
  ) |>
  hot_col(
    col = 3,
    type = "autocomplete",
    source = regions,
    strict = TRUE, # Only allow values from source
    filter = TRUE
  )
```

### Checkbox Columns

```{r checkbox-columns}
# Configure checkbox columns
task_data <- data.frame(
  Task = c("Review code", "Write tests", "Update docs"),
  Completed = c(TRUE, FALSE, FALSE),
  Priority = c("High", "Medium", "Low")
)

handsontable(task_data) |>
  hot_col(
    col = 2,
    type = "checkbox"
  )
```

### Date Columns

```{r date-columns}
# Configure date columns
schedule_data <- data.frame(
  Event = c("Meeting", "Deadline", "Review"),
  Date = as.Date(c("2024-01-15", "2024-01-20", "2024-01-25")),
  stringsAsFactors = FALSE
)

handsontable(schedule_data) |>
  hot_col(col = 2, type = "date", dateFormat = "YYYY-MM-DD")
```

### Password Columns

```{r password-columns}
# Configure password columns (text is hidden)
login_data <- data.frame(
  Username = c("user1", "user2", "admin"),
  Password = c("secret123", "pass456", "admin789"),
  Active = c(TRUE, TRUE, FALSE),
  stringsAsFactors = FALSE
)

handsontable(login_data) |>
  hot_col(
    col = 2,
    type = "password",
    copyable = FALSE # Prevent copying password values
  )
```

### Advanced Numeric Formatting

```{r advanced-numeric}
# Advanced numeric formatting options
financial_detailed <- data.frame(
  Item = c("Revenue", "Expenses", "Profit", "Tax"),
  Amount = c(1234567.89, -234567.12, 1000000.77, -150000.50),
  Percentage = c(0.15, -0.05, 0.12, -0.03),
  Units = c(1500000, 750000, 2250000, 500000)
)

handsontable(financial_detailed) |>
  hot_col(
    col = 2,
    type = "numeric",
    numericFormat = list(
      pattern = "$0,0.00",
      culture = "en-US"
    )
  ) |>
  hot_col(
    col = 3,
    type = "numeric",
    numericFormat = list(
      pattern = "0.00%"
    )
  ) |>
  hot_col(
    col = 4,
    type = "numeric",
    numericFormat = list(
      pattern = "0.0a" # Abbreviated format (1.5M, 750k, etc.)
    )
  )
```

## Individual Row Configuration with `hot_row()`

Configure specific rows with different properties.

### Read-Only Rows

```{r readonly-rows}
# Make specific rows read-only
data <- iris[1:8, ]

ht <- handsontable(data)
for (i in c(1, 3, 5)) {
  ht <- ht |> hot_row(row = i, readOnly = TRUE)
}
ht
```

## Context Menu Configuration with `hot_context_menu()`

Customize the right-click context menu.

### Basic Context Menu

```{r basic-context}
handsontable(mtcars[1:6, 1:4]) |>
  hot_context_menu()
```

### CSV Export (Built-in Feature)

The context menu automatically includes a "Download to CSV" option that properly handles empty values, NAs, and special characters:

```{r csv-export-basic}
# CSV export is automatically included when context menu is enabled
csv_test_data <- data.frame(
  Name = c("John", "Jane", "", "Bob", NA),
  Age = c(25, NA, 30, 35, 28),
  City = c("New York", "Boston", "", "Chicago", "Seattle"),
  Salary = c(50000, 60000, NA, 70000, 55000),
  Notes = c("Manager", "", "Developer, Senior", NA, "Designer"),
  Active = c(TRUE, FALSE, NA, TRUE, FALSE),
  stringsAsFactors = FALSE
)

# Right-click to see "Download to CSV" option
handsontable(csv_test_data) |>
  hot_context_menu()
```

### Custom Context Menu Configuration

```{r custom-context-menu}
# Use specific built-in context menu options
handsontable(csv_test_data[1:3, 1:3]) |>
  hot_context_menu(opts = c("row_above", "row_below", "---------", "copy", "cut"))
```

### Custom Actions with JavaScript Callbacks
```{r custom-context-callbacks}
# Add custom actions with JavaScript callbacks
custom_actions <- list(
  list(
    name = "Custom Action",
    callback = htmlwidgets::JS("function() { alert('Custom action clicked!'); }")
  )
)

handsontable(data = csv_test_data[1:5, 1:3]) |>
  hot_context_menu(
    opts = c("copy", "cut", "clear_column"),
    customOpts = custom_actions
  )
```

### Limited Context Menu Options
```{r limited-context}
# Only show specific context menu options
handsontable(iris[1:5, 1:3]) |>
  hot_context_menu(opts = c("copy", "undo", "redo"))
```

## Custom Borders with `customBorders`

Add custom borders around specific cell ranges to highlight important data.

### Basic Border Configuration

```{r basic-borders}
# Add red borders around specific cells
financial_data <- data.frame(
  Quarter = c("Q1", "Q2", "Q3", "Q4"),
  Revenue = c(120000, 135000, 148000, 162000),
  Profit = c(25000, 30000, 35000, 42000)
)

handsontable(financial_data) |>
  handsontable(
    customBorders = list(
      # Border around revenue header and data (0-based indexing)
      list(
        range = list(from = list(row = 0, col = 1), to = list(row = 3, col = 1)),
        top = list(width = 2, color = "blue"),
        right = list(width = 2, color = "blue"),
        bottom = list(width = 2, color = "blue"),
        left = list(width = 2, color = "blue")
      )
    )
  )
```

### Multiple Border Regions

```{r multiple-borders}
# Add different colored borders for different regions
performance_data <- data.frame(
  Team = c("Sales", "Marketing", "Engineering", "Support"),
  Q1 = c(95, 87, 92, 89),
  Q2 = c(88, 91, 94, 85),
  Q3 = c(92, 89, 96, 88),
  Q4 = c(97, 93, 98, 91)
)

handsontable(performance_data) |>
  handsontable(
    customBorders = list(
      # Green border for high performers (Engineering row - row 2, cols 1-4)
      list(
        range = list(from = list(row = 2, col = 1), to = list(row = 2, col = 4)),
        top = list(width = 3, color = "green"),
        right = list(width = 3, color = "green"),
        bottom = list(width = 3, color = "green"),
        left = list(width = 3, color = "green")
      ),
      # Red border for attention needed (Support Q2 - row 3, col 2)
      list(
        range = list(from = list(row = 3, col = 2), to = list(row = 3, col = 2)),
        top = list(width = 2, color = "red"),
        right = list(width = 2, color = "red"),
        bottom = list(width = 2, color = "red"),
        left = list(width = 2, color = "red")
      )
    )
  )
```

### Border Styles and Patterns

```{r border-styles}
# Different border styles for different purposes
budget_data <- data.frame(
  Category = c("Revenue", "Expenses", "Net Income", "Forecast"),
  Actual = c(250000, -180000, 70000, 75000),
  Budget = c(240000, -185000, 55000, 60000),
  Variance = c(10000, 5000, 15000, 15000)
)

handsontable(budget_data) |>
  handsontable(
    customBorders = list(
      # Thick blue border for Net Income row (row 2, all columns)
      list(
        range = list(from = list(row = 2, col = 0), to = list(row = 2, col = 3)),
        top = list(width = 3, color = "#2196F3"),
        bottom = list(width = 3, color = "#2196F3")
      ),
      # Orange border for forecast row (row 3, all columns)
      list(
        range = list(from = list(row = 3, col = 0), to = list(row = 3, col = 3)),
        top = list(width = 1, color = "#FFA726"),
        right = list(width = 1, color = "#FFA726"),
        bottom = list(width = 1, color = "#FFA726"),
        left = list(width = 1, color = "#FFA726")
      )
    )
  )
```

### Dynamic Border Updates

Borders can be updated dynamically in Shiny applications:

```{r dynamic-borders, eval=FALSE}
# Example for Shiny (not executed in vignette)
library(shiny)

ui <- fluidPage(
  handsontableOutput("dynamic_table"),
  actionButton("highlight_profit", "Highlight Profit Column"),
  actionButton("clear_borders", "Clear All Borders")
)

server <- function(input, output, session) {
  border_state <- reactiveVal(0)

  output$dynamic_table <- renderHandsontable({
    data <- data.frame(
      Product = c("A", "B", "C"),
      Sales = c(100, 200, 150),
      Profit = c(25, 50, 30)
    )

    if (border_state() == 1) {
      # Add borders around profit column (col 2, rows 0-2)
      handsontable(data) |>
        handsontable(
          customBorders = list(
            list(
              range = list(from = list(row = 0, col = 2), to = list(row = 2, col = 2)),
              top = list(width = 2, color = "gold"),
              right = list(width = 2, color = "gold"),
              bottom = list(width = 2, color = "gold"),
              left = list(width = 2, color = "gold")
            )
          )
        )
    } else {
      # No borders
      handsontable(data) |>
        handsontable(customBorders = list())
    }
  })

  observeEvent(input$highlight_profit, {
    border_state(1)
  })
  observeEvent(input$clear_borders, {
    border_state(0)
  })
}
```

### Border Configuration Notes

- **IMPORTANT: 0-based indexing**: Use 0-based indexing for borders (row 0 = first row, col 0 = first column)
- **Range Specification**: `from` and `to` define the cell range for borders
- **Border Sides**: Specify `top`, `right`, `bottom`, `left` independently
- **Width**: Border thickness in pixels (1-5 typically work well)
- **Color**: Use hex codes (#FF0000) or color names (red, blue, etc.)
- **Clearing Borders**: Set `customBorders = list()` to remove all borders
- **Performance**: Many borders can affect rendering performance on large tables
- **Indexing Note**: Unlike R's 1-based indexing, customBorders uses JavaScript's 0-based indexing

## Combining Multiple Configurations

Most real-world applications require combining multiple configuration options:

```{r complex-example}
# A comprehensive example with multiple configurations
sales_data <- data.frame(
  Date = seq(as.Date("2024-01-01"), by = "day", length.out = 10),
  Product = rep(c("Widget A", "Widget B"), 5),
  Quantity = sample(10:100, 10),
  Price = round(runif(10, 10, 50), 2),
  Total = 0, # Will be calculated
  Approved = rep(FALSE, 10)
)

handsontable(sales_data,
  filters = TRUE,
  manualColumnSorting = TRUE,
  search = TRUE,
  undo = TRUE
) |>
  # Configure individual columns
  hot_col(col = 1, type = "date", dateFormat = "YYYY-MM-DD", width = 100) |>
  hot_col(
    col = 2, type = "dropdown",
    source = c("Widget A", "Widget B", "Widget C"), width = 100
  ) |>
  hot_col(col = 3, type = "numeric", format = "0,0", width = 80) |>
  hot_col(col = 4, type = "numeric", format = "$0.00", width = 80) |>
  hot_col(col = 5, type = "numeric", format = "$0.00", readOnly = TRUE, width = 80) |>
  hot_col(col = 6, type = "checkbox", width = 80) |>
  # Configure context menu
  hot_context_menu()
```

## Tips for Configuration

### 1. Order Matters
Apply configurations in this typical order:
1. Column configurations (`hot_col()`)
1. Row configurations (`hot_row()`)
1. Table-wide features (`handsontable()`)
1. Context menu customizations (`hot_context_menu()`)

### 2. Column Indexing
- Columns are 1-indexed in R (first column is 1)
- You can use column names if `colHeaders = TRUE`

### 3. Performance Considerations
- For large datasets, consider limiting features like sorting and filtering
- Use `readOnly = TRUE` for display-only tables
- Set specific column widths to avoid auto-calculation overhead

### 4. Responsive Design
```{r responsive}
# Create responsive tables
handsontable(mtcars[1:5, 1:4], colWidths = "auto", stretchH = "all")
```

This covers the main configuration options available in the handsontable 
package. The next vignettes will cover data validation and Shiny integration.
