---
title: "Configuration Options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Configuration Options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = interactive()
)
```

```{r setup}
library(handsontable)
```

# Configuration Options

This vignette covers the various configuration functions available in the 
handsontable package, allowing you to customize every aspect of your tables.

## Column Configuration with `hot_cols()`

The `hot_cols()` function controls column-level settings that apply to all 
columns or specific ranges.

### Column Widths

```{r column-widths}
# Set uniform column width
handsontable(mtcars[1:5, 1:4]) |>
  hot_cols(colWidths = 120)

# Set individual column widths
handsontable(iris[1:5, ]) |>
  hot_cols(colWidths = c(80, 80, 80, 80, 120))
```

### Column Freezing and Resizing

```{r column-freezing}
# Freeze first 2 columns and enable column operations
handsontable(mtcars[1:8, ]) |>
  hot_cols(
    fixedColumnsLeft = 2,
    manualColumnResize = TRUE,
    manualColumnMove = TRUE
  )
```

## Row Configuration with `hot_rows()`

The `hot_rows()` function provides row-level configuration options.

### Row Freezing

```{r row-freezing}
# Freeze top 2 rows and bottom 1 row
handsontable(mtcars[1:10, 1:4]) |>
  hot_rows(
    fixedRowsTop = 2,
    fixedRowsBottom = 1,
    manualRowResize = TRUE,
    manualRowMove = TRUE
  )
```

## General Table Configuration with `hot_table()`

The `hot_table()` function controls overall table behavior and features.

### Context Menu and Interaction

```{r context-menu}
# Enable context menu and other interactive features
handsontable(iris[1:8, ]) |>
  hot_table(
    contextMenu = TRUE,
    filters = TRUE,
    manualColumnSorting = TRUE,
    search = TRUE,
    undo = TRUE
  )
```

### Multiple Column Sorting

```{r multi-sort}
# Enable multi-column sorting
handsontable(mtcars[1:10, 1:6]) |>
  hot_table(
    multiColumnSorting = TRUE,
    outsideClickDeselects = TRUE
  )
```

### Fill Handle Configuration

```{r fill-handle}
# Configure fill handle behavior
handsontable(data.frame(
  Numbers = 1:5,
  Letters = letters[1:5]
)) |>
  hot_table(
    fillHandle = list(
      direction = "vertical",
      autoInsertRow = TRUE
    )
  )
```

## Individual Column Configuration with `hot_col()`

Use `hot_col()` to configure specific columns with different types and 
behaviors.

### Text Columns

```{r text-columns}
# Configure a text column
data <- data.frame(
  ID = 1:5,
  Name = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  Notes = rep("", 5)
)

handsontable(data) |>
  hot_col(col = 1, readOnly = TRUE, width = 50) |>
  hot_col(col = 2, type = "text", width = 100) |>
  hot_col(col = 3, type = "text", width = 200)
```

### Numeric Columns

```{r numeric-columns}
# Configure numeric columns with formatting
financial_data <- data.frame(
  Product = c("Widget A", "Widget B", "Widget C"),
  Price = c(19.99, 24.50, 15.75),
  Quantity = c(100, 250, 75)
)

handsontable(financial_data) |>
  hot_col(col = 2, type = "numeric", format = "$0.00") |>
  hot_col(col = 3, type = "numeric", format = "0,0")
```

### Dropdown Columns

```{r dropdown-columns}
# Create dropdown columns
employee_data <- data.frame(
  Name = c("John", "Jane", "Bob"),
  Department = c("IT", "HR", "Finance"),
  Level = c("Senior", "Junior", "Manager"),
  stringsAsFactors = FALSE
)

handsontable(employee_data) |>
  hot_col(
    col = 2,
    type = "dropdown",
    source = c("IT", "HR", "Finance", "Marketing", "Sales"),
    strict = TRUE
  ) |>
  hot_col(
    col = 3,
    type = "dropdown",
    source = c("Junior", "Senior", "Manager", "Director"),
    strict = TRUE
  )
```

### Autocomplete Columns

```{r autocomplete-columns}
# Create autocomplete columns (like dropdown but with filtering)
country_data <- data.frame(
  City = c("New York", "London", "Tokyo"),
  Country = c("USA", "UK", "Japan"),
  Region = c("North America", "Europe", "Asia"),
  stringsAsFactors = FALSE
)

countries <- c("USA", "UK", "Japan", "Germany", "France", "Canada", "Australia", "Brazil")
regions <- c("North America", "Europe", "Asia", "Oceania", "South America")

handsontable(country_data) |>
  hot_col(
    col = 2,
    type = "autocomplete",
    source = countries,
    strict = FALSE, # Allow new values
    filter = TRUE, # Enable filtering as you type
    trimDropdown = TRUE # Fit dropdown to content
  ) |>
  hot_col(
    col = 3,
    type = "autocomplete",
    source = regions,
    strict = TRUE, # Only allow values from source
    filter = TRUE
  )
```

### Checkbox Columns

```{r checkbox-columns}
# Configure checkbox columns
task_data <- data.frame(
  Task = c("Review code", "Write tests", "Update docs"),
  Completed = c(TRUE, FALSE, FALSE),
  Priority = c("High", "Medium", "Low")
)

handsontable(task_data) |>
  hot_col(
    col = 2,
    type = "checkbox"
  )
```

### Date Columns

```{r date-columns}
# Configure date columns
schedule_data <- data.frame(
  Event = c("Meeting", "Deadline", "Review"),
  Date = as.Date(c("2024-01-15", "2024-01-20", "2024-01-25")),
  stringsAsFactors = FALSE
)

handsontable(schedule_data) |>
  hot_col(col = 2, type = "date", dateFormat = "YYYY-MM-DD")
```

### Password Columns

```{r password-columns}
# Configure password columns (text is hidden)
login_data <- data.frame(
  Username = c("user1", "user2", "admin"),
  Password = c("secret123", "pass456", "admin789"),
  Active = c(TRUE, TRUE, FALSE),
  stringsAsFactors = FALSE
)

handsontable(login_data) |>
  hot_col(
    col = 2,
    type = "password",
    copyable = FALSE # Prevent copying password values
  )
```

### Advanced Numeric Formatting

```{r advanced-numeric}
# Advanced numeric formatting options
financial_detailed <- data.frame(
  Item = c("Revenue", "Expenses", "Profit", "Tax"),
  Amount = c(1234567.89, -234567.12, 1000000.77, -150000.50),
  Percentage = c(0.15, -0.05, 0.12, -0.03),
  Units = c(1500000, 750000, 2250000, 500000)
)

handsontable(financial_detailed) |>
  hot_col(
    col = 2,
    type = "numeric",
    numericFormat = list(
      pattern = "$0,0.00",
      culture = "en-US"
    )
  ) |>
  hot_col(
    col = 3,
    type = "numeric",
    numericFormat = list(
      pattern = "0.00%"
    )
  ) |>
  hot_col(
    col = 4,
    type = "numeric",
    numericFormat = list(
      pattern = "0.0a" # Abbreviated format (1.5M, 750k, etc.)
    )
  )
```

## Individual Row Configuration with `hot_row()`

Configure specific rows with different properties.

### Read-Only Rows

```{r readonly-rows}
# Make specific rows read-only
data <- iris[1:8, ]

ht <- handsontable(data)
for (i in c(1, 3, 5)) {
  ht <- ht |> hot_row(row = i, readOnly = TRUE)
}
ht
```

## Context Menu Configuration with `hot_context_menu()`

Customize the right-click context menu.

### Basic Context Menu

```{r basic-context}
handsontable(mtcars[1:6, 1:4]) |>
  hot_table(contextMenu = TRUE) |>
  hot_context_menu(
    allowRowEdit = TRUE,
    allowColEdit = TRUE
  )
```

### CSV Export (Built-in Feature)

The context menu automatically includes a "Download to CSV" option that properly handles empty values, NAs, and special characters:

```{r csv-export-basic}
# CSV export is automatically included when context menu is enabled
csv_test_data <- data.frame(
  Name = c("John", "Jane", "", "Bob", NA),
  Age = c(25, NA, 30, 35, 28),
  City = c("New York", "Boston", "", "Chicago", "Seattle"),
  Salary = c(50000, 60000, NA, 70000, 55000),
  Notes = c("Manager", "", "Developer, Senior", NA, "Designer"),
  Active = c(TRUE, FALSE, NA, TRUE, FALSE),
  stringsAsFactors = FALSE
)

# Right-click to see "Download to CSV" option
handsontable(csv_test_data) |>
  hot_table(contextMenu = TRUE)
```

### Custom Context Menu Options

#### Built-in Actions (Recommended)
```{r custom-context-builtin}
# Use built-in Handsontable actions for full functionality
# CSV export is automatically included
builtin_options <- list(
  "copy" = TRUE, # Built-in copy
  "paste" = TRUE, # Built-in paste
  "sep1" = "---------", # Separator
  "clear_column" = TRUE, # Built-in clear column
  "remove_row" = TRUE # Built-in remove row
)

handsontable(iris[1:5, 1:3]) |>
  hot_context_menu(customOpts = builtin_options)
```

#### Custom Actions with JavaScript Callbacks
```{r custom-context-callbacks}
# Add custom actions with JavaScript callbacks
# CSV export is still included automatically
custom_actions <- list(
  "copy" = TRUE,
  "cut" = TRUE,
  "separator1" = "---------",
  custom_action = list(
    name = "Custom Action",
    callback = htmlwidgets::JS("function() { alert('Custom action clicked!'); }")
  )
)

handsontable(csv_test_data[1:5, 1:3]) |>
  hot_context_menu(customOpts = custom_actions)
```

#### Disabling CSV Export
```{r disable-csv-export}
# Users can disable CSV export if they don't want it
no_csv_options <- list(
  "export_csv" = FALSE, # This disables the CSV export
  "copy" = TRUE,
  "cut" = TRUE,
  "separator1" = "---------",
  "remove_row" = TRUE
)

handsontable(iris[1:5, 1:3]) |>
  hot_context_menu(customOpts = no_csv_options)
```

#### Custom Labels with Built-in Actions
```{r custom-context-labels}
# Customize labels while keeping built-in functionality
custom_labels <- list(
  "copy" = list(name = "Copy cells"),
  "paste" = list(name = "Paste cells"),
  "sep1" = "---------",
  "clear_column" = list(name = "Clear column")
  # CSV export keeps default label "Download to CSV"
)

handsontable(iris[1:5, 1:3]) |>
  hot_context_menu(customOpts = custom_labels)
```

## Combining Multiple Configurations

Most real-world applications require combining multiple configuration options:

```{r complex-example}
# A comprehensive example with multiple configurations
sales_data <- data.frame(
  Date = seq(as.Date("2024-01-01"), by = "day", length.out = 10),
  Product = rep(c("Widget A", "Widget B"), 5),
  Quantity = sample(10:100, 10),
  Price = round(runif(10, 10, 50), 2),
  Total = 0, # Will be calculated
  Approved = rep(FALSE, 10)
)

handsontable(sales_data) |>
  # Configure individual columns
  hot_col(col = 1, type = "date", dateFormat = "YYYY-MM-DD", width = 100) |>
  hot_col(
    col = 2, type = "dropdown",
    source = c("Widget A", "Widget B", "Widget C"), width = 100
  ) |>
  hot_col(col = 3, type = "numeric", format = "0,0", width = 80) |>
  hot_col(col = 4, type = "numeric", format = "$0.00", width = 80) |>
  hot_col(col = 5, type = "numeric", format = "$0.00", readOnly = TRUE, width = 80) |>
  hot_col(col = 6, type = "checkbox", width = 80) |>
  # Configure table features
  hot_table(
    contextMenu = TRUE,
    filters = TRUE,
    manualColumnSorting = TRUE,
    search = TRUE,
    undo = TRUE
  ) |>
  # Configure columns and rows
  hot_cols(
    fixedColumnsLeft = 1,
    manualColumnResize = TRUE
  ) |>
  hot_rows(
    manualRowResize = TRUE,
    manualRowMove = TRUE
  )
```

## Tips for Configuration

### 1. Order Matters
Apply configurations in this typical order:
1. Individual column configurations (`hot_col()`)
2. Individual row configurations (`hot_row()`)
3. Column-wide settings (`hot_cols()`)
4. Row-wide settings (`hot_rows()`)
5. Table-wide features (`hot_table()`)
6. Context menu customizations (`hot_context_menu()`)

### 2. Column Indexing
- Columns are 1-indexed in R (first column is 1)
- You can use column names if `colHeaders = TRUE`

### 3. Performance Considerations
- For large datasets, consider limiting features like sorting and filtering
- Use `readOnly = TRUE` for display-only tables
- Set specific column widths to avoid auto-calculation overhead

### 4. Responsive Design
```{r responsive}
# Create responsive tables
handsontable(mtcars[1:5, 1:4]) |>
  hot_cols(colWidths = "auto") |> # Auto-size columns
  hot_table(stretchH = "all") # Stretch to container
```

This covers the main configuration options available in the handsontable 
package. The next vignettes will cover data validation and Shiny integration.
