---
title: "Getting Started with handsontable"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with handsontable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = interactive()
)
```

```{r setup}
library(handsontable)
```

# Introduction

The `handsontable` package provides an R interface to the powerful Handsontable 
JavaScript library, bringing spreadsheet-like functionality to R and Shiny 
applications. This vignette will get you started with the basics.

## What is Handsontable?

Handsontable is an interactive data grid that combines the familiarity of 
spreadsheets with the power of professional data grids. It provides:

- **Interactive editing** - Click to edit cells directly
- **Data validation** - Ensure data quality with built-in validators
- **Flexible formatting** - Control appearance and behavior
- **Shiny integration** - Seamless reactive programming support

## Your First Handsontable

The simplest way to create a handsontable is to pass a data.frame to the 
`handsontable()` function:

```{r basic-example}
# Create a basic handsontable
handsontable(mtcars[1:5, 1:4])
```

This creates an interactive table with:
- Column headers from the data.frame column names
- Row numbers on the left
- Editable cells
- Automatic column width adjustment

## Working with Different Data Types

Handsontable automatically handles different R data types:

```{r data-types}
# Create a dataset with mixed types
mixed_data <- data.frame(
  Name = c("Alice", "Bob", "Charlie", "Diana"),
  Age = c(25, 30, 35, 28),
  Salary = c(50000, 65000, 75000, 60000),
  Active = c(TRUE, FALSE, TRUE, TRUE),
  Department = factor(c("IT", "HR", "Finance", "IT")),
  Start_Date = as.Date(c("2020-01-15", "2019-03-10", "2018-07-22", "2021-02-01"))
)

handsontable(mixed_data)
```

Notice how:

- Text is displayed as-is
- Numbers are right-aligned
- Logical values show as TRUE/FALSE
- Factors are converted to character strings
- Dates are displayed in a readable format

## Basic Customization

### Custom Headers

You can provide custom column headers:

```{r custom-headers}
handsontable(
  iris[1:8, ],
  colHeaders = c("Sepal L", "Sepal W", "Petal L", "Petal W", "Species")
)
```

### Controlling Headers and Sizing

```{r headers-sizing}
handsontable(
  mtcars[, 1:4],
  colHeaders = TRUE, # Show column headers (default)
  rowHeaders = FALSE, # Hide row numbers
  width = "100%", # Full width
  height = "300px", # Fixed height
  adaptiveHeight = FALSE
)
```

### Read-Only Tables

Make the entire table read-only:

```{r readonly}
handsontable(
  iris[1:5, ],
  readOnly = TRUE
)
```

## Matrix Input

Handsontable also works with matrices:

```{r matrix-input}
# Create a matrix
mat <- matrix(1:12, nrow = 3, ncol = 4)
colnames(mat) <- paste("Col", 1:4)
rownames(mat) <- paste("Row", 1:3)

handsontable(mat)
```

## Understanding the Widget Output

When you create a handsontable, you get an htmlwidget object that contains:

```{r widget-structure, eval=FALSE}
ht <- handsontable(iris[1:3, 1:3])

# The widget has these main components:
# - x$data: The data in JavaScript format
# - x$colHeaders: Column header names
# - x$rowHeaders: Whether to show row headers
# - x$readOnly: Whether the table is read-only
```

## Basic Pipe Operations

The package is designed to work well with the pipe operator:

```{r pipe-example}
iris[1:5, ] |>
  handsontable() |>
  hot_cols(colWidths = 100) |>
  hot_table(contextMenu = TRUE)
```

## Next Steps

This introduction covered the basics of creating handsontables. In the next 
vignettes, you'll learn about:

- **Configuration Options**: Detailed customization of appearance and behavior
- **Data Validation**: Ensuring data quality with validators
- **Shiny Integration**: Building interactive applications
- **Advanced Features**: Context menus, sorting, filtering, and more

Try experimenting with the examples above and see how the tables respond to 
your interactions!
