---
title: "Shiny Integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shiny Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE # Set to FALSE since these are Shiny examples
)
```

```{r setup}
library(handsontable)
library(shiny)
```

# Shiny Integration

The handsontable package is designed for seamless integration with Shiny 
applications. This vignette covers everything you need to know about using 
handsontables in reactive applications.

## Basic Shiny Integration

### Essential Components

Every Shiny app with handsontables needs three main components:

1. **Output function**: `handsontableOutput()` in the UI
2. **Render function**: `renderHandsontable()` in the server
3. **Input detection**: Reactive access to table changes

### Minimal Example

```{r minimal-shiny}
library(shiny)
library(handsontable)

ui <- fluidPage(
  titlePanel("Basic Handsontable"),
  handsontableOutput("table")
)

server <- function(input, output, session) {
  output$table <- renderHandsontable({
    handsontable(mtcars[1:6, 1:4])
  })
}

shinyApp(ui, server)
```

## Detecting Changes

The most powerful feature of Shiny integration is the ability to react to 
table changes in real-time.

### Basic Change Detection

```{r change-detection}
library(shiny)
library(handsontable)

ui <- fluidPage(
  titlePanel("Change Detection"),
  handsontableOutput("table"),
  br(),
  h4("Change Information:"),
  verbatimTextOutput("change_info")
)

server <- function(input, output, session) {
  # Create initial data
  initial_data <- data.frame(
    Name = c("Alice", "Bob", "Charlie"),
    Age = c(25, 30, 35),
    Department = c("IT", "HR", "Finance")
  )

  output$table <- renderHandsontable({
    handsontable(initial_data)
  })

  # React to any change in the table
  observeEvent(input$table, {
    cat("Table changed at:", as.character(Sys.time()), "\n")

    if (!is.null(input$table)) {
      # Get the current data
      current_data <- hot_to_r(input$table)
      print(current_data)

      # Check what type of change occurred
      if (!is.null(input$table$changes)) {
        cat("Number of cell changes:", length(input$table$changes), "\n")
      }
    }
  })

  output$change_info <- renderText({
    if (is.null(input$table)) {
      "No changes yet"
    } else {
      paste(
        "Last change at:",
        Sys.time(),
        "\n",
        "Event type:",
        input$table$event %||% "afterChange",
        "\n",
        "Data rows:",
        length(input$table$data %||% list())
      )
    }
  })
}

shinyApp(ui, server)
```

### Available Input Values

When a handsontable changes, several input values become available:

- **`input$table`** - Main input with complete information
- `input$table$event`: The event hook that triggered the change:
  - "afterChange"
  - "afterSelection"
  - "afterCreateRow"
  - "afteRemoveRow"
  - "afterCreateCol"
  - "afterRemoveCol"
- **`input$table_data`** - Just the current data
- **`input$table_changes`** - Specific change details
- **`input$table_select`** - Current selection information

## Event Hooks

Handsontable provides several event hooks that allow you to react to 
specific user interactions. Each hook sends detailed information about 
the event, including an `event` type identifier and additional context-specific parameters.

### afterChange Hook

The `afterChange` hook fires whenever cell data is modified. It provides information about what changed, including the source of the change (edit, paste, fill, etc.).

<details>
<summary><strong>Click to expand afterChange example</strong></summary>

```{r afterchange-hook}
library(shiny)
library(handsontable)

# Sample data
sample_data <- data.frame(
  ID = 1:5,
  Name = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  Score = c(85, 92, 78, 95, 88),
  Grade = c("B", "A", "C", "A", "B"),
  Active = c(TRUE, FALSE, TRUE, TRUE, FALSE)
)

ui <- fluidPage(
  titlePanel("afterChange Hook Demo - Event & Source Parameters"),
  
  fluidRow(
    column(8,
      h4("Interactive Handsontable"),
      p("Try editing cells, pasting data, or using the context menu to see different source types:"),
      tags$ul(
        tags$li("Edit a cell directly (source: 'edit')"),
        tags$li("Copy/paste data (source: 'paste')"),
        tags$li("Use fill handle to drag data (source: 'autofill')"),
        tags$li("Right-click and insert/remove rows (source: 'UndoRedo.redo' or similar)")
      ),
      handsontableOutput("demo_table", height = "300px")
    ),
    
    column(4,
      h4("Change Event Log"),
      p("Real-time log of afterChange events:"),
      verbatimTextOutput("change_log", placeholder = TRUE),
      
      br(),
      h4("Current Data Structure"),
      verbatimTextOutput("data_structure", placeholder = TRUE),
      
      br(),
      actionButton("clear_log", "Clear Log", class = "btn-warning btn-sm"),
      br(), br(),
      actionButton("programmatic_update", "Programmatic Update", class = "btn-info btn-sm"),
      helpText("Updates Score column programmatically to test different source types")
    )
  )
)

server <- function(input, output, session) {
  # Reactive values to store change log
  change_log <- reactiveVal(character(0))
  
  # Render the handsontable
  output$demo_table <- renderHandsontable({
    handsontable(sample_data) |>
      hot_col(col = 1, readOnly = TRUE, width = 60) |>
      hot_col(col = 3, type = "numeric", format = "0") |>
      hot_col(col = 4, type = "dropdown", source = c("A", "B", "C", "D", "F")) |>
      hot_col(col = 5, type = "checkbox") |>
      hot_table(contextMenu = TRUE, undo = TRUE)
  })
  
  # Observe changes in the handsontable
  observeEvent(input$demo_table, {
    if (!is.null(input$demo_table)) {
      # Check if this is an afterChange event
      if (!is.null(input$demo_table$event) && input$demo_table$event == "afterChange") {
        
        # Extract event details
        event_type <- input$demo_table$event
        source_type <- input$demo_table$source %||% "unknown"
        changes <- input$demo_table$changes
        
        # Format the change information
        if (!is.null(changes) && length(changes) > 0) {
          # Format timestamp
          timestamp <- format(Sys.time(), "%H:%M:%S")
          
          # Format changes - each change is [row, col, oldValue, newValue]
          change_details <- sapply(changes, function(change) {
            row_idx <- change[[1]] + 1  # Convert 0-based to 1-based indexing
            col_idx <- change[[2]] + 1
            old_val <- if(is.null(change[[3]])) "NULL" else as.character(change[[3]])
            new_val <- if(is.null(change[[4]])) "NULL" else as.character(change[[4]])
            
            sprintf("  Row %d, Col %d: '%s' → '%s'", row_idx, col_idx, old_val, new_val)
          })
          
          # Create log entry
          log_entry <- paste0(
            "[", timestamp, "] EVENT: ", event_type, " | SOURCE: ", source_type, "\n",
            paste(change_details, collapse = "\n")
          )
          
          # Add to log (keep last 10 entries)
          current_log <- change_log()
          new_log <- c(log_entry, current_log)
          if (length(new_log) > 10) {
            new_log <- new_log[1:10]
          }
          change_log(new_log)
        }
      }
    }
  })
  
  # Display the change log
  output$change_log <- renderText({
    log_entries <- change_log()
    if (length(log_entries) == 0) {
      "No changes yet. Try editing some cells!"
    } else {
      paste(log_entries, collapse = "\n\n---\n\n")
    }
  })
  
  # Display current data structure
  output$data_structure <- renderText({
    if (!is.null(input$demo_table) && !is.null(input$demo_table$data)) {
      current_data <- hot_to_r(input$demo_table)
      paste(
        "Dimensions:", nrow(current_data), "rows ×", ncol(current_data), "columns\n",
        "Column names:", paste(names(current_data), collapse = ", "), "\n",
        "Last updated:", format(Sys.time(), "%H:%M:%S")
      )
    } else {
      "No data available yet."
    }
  })
  
  # Clear log button
  observeEvent(input$clear_log, {
    change_log(character(0))
  })
  
  # Programmatic update button - demonstrates different source types
  observeEvent(input$programmatic_update, {
    # Update some cells programmatically using set_data
    new_scores <- sample(70:100, 2)
    set_data("demo_table", row = c(2, 4), col = 3, val = new_scores)
    
    showNotification(
      paste("Updated scores for rows 2 and 4 to:", paste(new_scores, collapse = ", ")),
      type = "message",
      duration = 3
    )
  })
}

shinyApp(ui, server)
```

</details>

### afterSelection Hook

The `afterSelection` hook fires when users select cells or ranges in the table.

<details>
<summary><strong>Click to expand afterSelection example</strong></summary>

```{r afterselection-hook}
library(shiny)
library(handsontable)

ui <- fluidPage(
  titlePanel("afterSelection Hook Demo"),
  
  fluidRow(
    column(8,
      h4("Select cells to see selection details"),
      p("Try selecting single cells, ranges, or multiple selections:"),
      handsontableOutput("selection_table", height = "300px")
    ),
    
    column(4,
      h4("Selection Information"),
      verbatimTextOutput("selection_info"),
      
      br(),
      h4("Selected Data"),
      p("Data from currently selected cells:"),
      tableOutput("selected_data")
    )
  )
)

server <- function(input, output, session) {
  # Sample data
  table_data <- data.frame(
    Product = c("Widget A", "Widget B", "Widget C", "Widget D", "Widget E"),
    Category = c("Electronics", "Toys", "Electronics", "Books", "Toys"),
    Price = c(25.99, 12.50, 45.00, 18.99, 8.75),
    Stock = c(100, 250, 75, 180, 320),
    Supplier = c("Acme Corp", "Toy Co", "Acme Corp", "Book Inc", "Toy Co")
  )
  
  output$selection_table <- renderHandsontable({
    handsontable(table_data) |>
      hot_col(col = 3, type = "numeric", format = "$0.00") |>
      hot_col(col = 4, type = "numeric", format = "0,0") |>
      hot_table(contextMenu = TRUE)
  })
  
  # Display selection information
  output$selection_info <- renderText({
    if (is.null(input$selection_table_select)) {
      "No selection made yet"
    } else {
      sel <- input$selection_table_select
      
      # Convert 0-based to 1-based indexing
      start_row <- sel$row + 1
      start_col <- sel$col + 1
      end_row <- sel$row2 + 1
      end_col <- sel$col2 + 1
      
      if (start_row == end_row && start_col == end_col) {
        paste0(
          "Single cell selected:\n",
          "Row: ", start_row, ", Column: ", start_col, "\n",
          "Value: ", table_data[start_row, start_col]
        )
      } else {
        paste0(
          "Range selected:\n",
          "From Row ", start_row, ", Col ", start_col, "\n",
          "To Row ", end_row, ", Col ", end_col, "\n",
          "Total cells: ", (end_row - start_row + 1) * (end_col - start_col + 1)
        )
      }
    }
  })
  
  # Display selected data
  output$selected_data <- renderTable({
    if (is.null(input$selection_table_select)) {
      data.frame(Message = "No selection")
    } else {
      sel <- input$selection_table_select
      
      # Convert 0-based to 1-based indexing
      start_row <- sel$row + 1
      start_col <- sel$col + 1
      end_row <- sel$row2 + 1
      end_col <- sel$col2 + 1
      
      # Extract selected data
      selected_data <- table_data[start_row:end_row, start_col:end_col, drop = FALSE]
      selected_data
    }
  })
}

shinyApp(ui, server)
```

</details>

### afterCreateRow Hook

The `afterCreateRow` hook fires when new rows are added to the table.

<details>
<summary><strong>Click to expand afterCreateRow example</strong></summary>

```{r aftercreaterow-hook}
library(shiny)
library(handsontable)

ui <- fluidPage(
  titlePanel("afterCreateRow Hook Demo"),
  
  fluidRow(
    column(8,
      h4("Right-click and insert rows to see the hook in action"),
      p("Use the context menu to add new rows:"),
      handsontableOutput("create_row_table", height = "300px")
    ),
    
    column(4,
      h4("Row Creation Log"),
      verbatimTextOutput("row_create_log"),
      
      br(),
      actionButton("clear_row_log", "Clear Log", class = "btn-warning btn-sm"),
      
      br(), br(),
      h4("Current Row Count"),
      textOutput("row_count")
    )
  )
)

server <- function(input, output, session) {
  # Initial data
  initial_data <- data.frame(
    Task = c("Task 1", "Task 2", "Task 3"),
    Status = c("Pending", "In Progress", "Completed"),
    Priority = c("High", "Medium", "Low"),
    Assigned = c("Alice", "Bob", "Charlie")
  )
  
  # Reactive values
  row_create_log <- reactiveVal(character(0))
  
  output$create_row_table <- renderHandsontable({
    handsontable(initial_data) |>
      hot_col(col = 2, type = "dropdown", source = c("Pending", "In Progress", "Completed")) |>
      hot_col(col = 3, type = "dropdown", source = c("High", "Medium", "Low")) |>
      hot_table(contextMenu = TRUE)
  })
  
  # Observe table changes
  observeEvent(input$create_row_table, {
    if (!is.null(input$create_row_table)) {
      # Check if this is an afterCreateRow event
      if (!is.null(input$create_row_table$event) && 
          input$create_row_table$event == "afterCreateRow") {
        
        # Extract event details
        index <- input$create_row_table$index + 1  # Convert to 1-based
        amount <- input$create_row_table$amount
        
        # Create log entry
        timestamp <- format(Sys.time(), "%H:%M:%S")
        log_entry <- paste0(
          "[", timestamp, "] EVENT: afterCreateRow\n",
          "  Inserted ", amount, " row(s) at position ", index
        )
        
        # Add to log
        current_log <- row_create_log()
        new_log <- c(log_entry, current_log)
        if (length(new_log) > 10) {
          new_log <- new_log[1:10]
        }
        row_create_log(new_log)
      }
    }
  })
  
  # Display creation log
  output$row_create_log <- renderText({
    log_entries <- row_create_log()
    if (length(log_entries) == 0) {
      "No rows created yet. Right-click and select 'Insert row above/below'."
    } else {
      paste(log_entries, collapse = "\n\n---\n\n")
    }
  })
  
  # Clear log button
  observeEvent(input$clear_row_log, {
    row_create_log(character(0))
  })
  
  # Display current row count
  output$row_count <- renderText({
    if (!is.null(input$create_row_table) && !is.null(input$create_row_table$data)) {
      paste("Current rows:", length(input$create_row_table$data))
    } else {
      paste("Current rows:", nrow(initial_data))
    }
  })
}

shinyApp(ui, server)
```

</details>

### afterRemoveRow Hook

The `afterRemoveRow` hook fires when rows are deleted from the table.

<details>
<summary><strong>Click to expand afterRemoveRow example</strong></summary>

```{r afterremoverow-hook}
library(shiny)
library(handsontable)

ui <- fluidPage(
  titlePanel("afterRemoveRow Hook Demo"),
  
  fluidRow(
    column(8,
      h4("Right-click and remove rows to see the hook in action"),
      p("Use the context menu to delete rows:"),
      handsontableOutput("remove_row_table", height = "300px")
    ),
    
    column(4,
      h4("Row Deletion Log"),
      verbatimTextOutput("row_delete_log"),
      
      br(),
      actionButton("clear_delete_log", "Clear Log", class = "btn-warning btn-sm"),
      
      br(), br(),
      h4("Deletion Summary"),
      verbatimTextOutput("delete_summary")
    )
  )
)

server <- function(input, output, session) {
  # Initial data with more rows for deletion testing
  initial_data <- data.frame(
    ID = 1:8,
    Employee = paste("Employee", 1:8),
    Department = sample(c("IT", "HR", "Finance", "Marketing"), 8, replace = TRUE),
    Salary = sample(40000:80000, 8),
    Active = sample(c(TRUE, FALSE), 8, replace = TRUE)
  )
  
  # Reactive values
  row_delete_log <- reactiveVal(character(0))
  delete_count <- reactiveVal(0)
  
  output$remove_row_table <- renderHandsontable({
    handsontable(initial_data) |>
      hot_col(col = 1, readOnly = TRUE) |>
      hot_col(col = 3, type = "dropdown", source = c("IT", "HR", "Finance", "Marketing")) |>
      hot_col(col = 4, type = "numeric", format = "$0,0") |>
      hot_col(col = 5, type = "checkbox") |>
      hot_table(contextMenu = TRUE)
  })
  
  # Observe table changes
  observeEvent(input$remove_row_table, {
    if (!is.null(input$remove_row_table)) {
      # Check if this is an afterRemoveRow event
      if (!is.null(input$remove_row_table$event) && 
          input$remove_row_table$event == "afterRemoveRow") {
        
        # Extract event details
        index <- input$remove_row_table$index + 1  # Convert to 1-based
        amount <- input$remove_row_table$amount
        
        # Update delete count
        delete_count(delete_count() + amount)
        
        # Create log entry
        timestamp <- format(Sys.time(), "%H:%M:%S")
        log_entry <- paste0(
          "[", timestamp, "] EVENT: afterRemoveRow\n",
          "  Removed ", amount, " row(s) starting at position ", index, "\n",
          "  Total rows remaining: ", length(input$remove_row_table$data)
        )
        
        # Add to log
        current_log <- row_delete_log()
        new_log <- c(log_entry, current_log)
        if (length(new_log) > 10) {
          new_log <- new_log[1:10]
        }
        row_delete_log(new_log)
      }
    }
  })
  
  # Display deletion log
  output$row_delete_log <- renderText({
    log_entries <- row_delete_log()
    if (length(log_entries) == 0) {
      "No rows deleted yet. Right-click and select 'Delete row'."
    } else {
      paste(log_entries, collapse = "\n\n---\n\n")
    }
  })
  
  # Clear log button
  observeEvent(input$clear_delete_log, {
    row_delete_log(character(0))
    delete_count(0)
  })
  
  # Display deletion summary
  output$delete_summary <- renderText({
    current_rows <- if (!is.null(input$remove_row_table$data)) {
      length(input$remove_row_table$data)
    } else {
      nrow(initial_data)
    }
    
    paste0(
      "Original rows: ", nrow(initial_data), "\n",
      "Current rows: ", current_rows, "\n",
      "Total deleted: ", delete_count(), "\n",
      "Remaining: ", current_rows
    )
  })
}

shinyApp(ui, server)
```

</details>

### afterCreateCol Hook

The `afterCreateCol` hook fires when new columns are added to the table.

<details>
<summary><strong>Click to expand afterCreateCol example</strong></summary>

```{r aftercreatecol-hook}
library(shiny)
library(handsontable)

ui <- fluidPage(
  titlePanel("afterCreateCol Hook Demo"),
  
  fluidRow(
    column(8,
      h4("Right-click and insert columns to see the hook in action"),
      p("Use the context menu to add new columns:"),
      handsontableOutput("create_col_table", height = "300px")
    ),
    
    column(4,
      h4("Column Creation Log"),
      verbatimTextOutput("col_create_log"),
      
      br(),
      actionButton("clear_col_log", "Clear Log", class = "btn-warning btn-sm"),
      
      br(), br(),
      h4("Column Information"),
      verbatimTextOutput("col_info")
    )
  )
)

server <- function(input, output, session) {
  # Initial data
  initial_data <- data.frame(
    Name = c("Alice", "Bob", "Charlie"),
    Age = c(25, 30, 35),
    City = c("New York", "London", "Tokyo")
  )
  
  # Reactive values
  col_create_log <- reactiveVal(character(0))
  
  output$create_col_table <- renderHandsontable({
    handsontable(initial_data) |>
      hot_col(col = 2, type = "numeric") |>
      hot_table(contextMenu = TRUE)
  })
  
  # Observe table changes
  observeEvent(input$create_col_table, {
    if (!is.null(input$create_col_table)) {
      # Check if this is an afterCreateCol event
      if (!is.null(input$create_col_table$event) && 
          input$create_col_table$event == "afterCreateCol") {
        
        # Extract event details
        index <- input$create_col_table$index + 1  # Convert to 1-based
        amount <- input$create_col_table$amount
        colnames <- input$create_col_table$colnames
        
        # Create log entry
        timestamp <- format(Sys.time(), "%H:%M:%S")
        log_entry <- paste0(
          "[", timestamp, "] EVENT: afterCreateCol\n",
          "  Inserted ", amount, " column(s) at position ", index, "\n",
          "  New column names: ", paste(colnames[index:(index + amount - 1)], collapse = ", ")
        )
        
        # Add to log
        current_log <- col_create_log()
        new_log <- c(log_entry, current_log)
        if (length(new_log) > 10) {
          new_log <- new_log[1:10]
        }
        col_create_log(new_log)
      }
    }
  })
  
  # Display creation log
  output$col_create_log <- renderText({
    log_entries <- col_create_log()
    if (length(log_entries) == 0) {
      "No columns created yet. Right-click and select 'Insert column left/right'."
    } else {
      paste(log_entries, collapse = "\n\n---\n\n")
    }
  })
  
  # Clear log button
  observeEvent(input$clear_col_log, {
    col_create_log(character(0))
  })
  
  # Display column information
  output$col_info <- renderText({
    if (!is.null(input$create_col_table) && !is.null(input$create_col_table$colnames)) {
      colnames <- input$create_col_table$colnames
      paste0(
        "Current columns: ", length(colnames), "\n",
        "Column names: ", paste(colnames, collapse = ", ")
      )
    } else {
      paste0(
        "Current columns: ", ncol(initial_data), "\n",
        "Column names: ", paste(names(initial_data), collapse = ", ")
      )
    }
  })
}

shinyApp(ui, server)
```

</details>

### afterRemoveCol Hook

The `afterRemoveCol` hook fires when columns are deleted from the table.

<details>
<summary><strong>Click to expand afterRemoveCol example</strong></summary>

```{r afterremovecol-hook}
library(shiny)
library(handsontable)

ui <- fluidPage(
  titlePanel("afterRemoveCol Hook Demo"),
  
  fluidRow(
    column(8,
      h4("Right-click and remove columns to see the hook in action"),
      p("Use the context menu to delete columns:"),
      handsontableOutput("remove_col_table", height = "300px")
    ),
    
    column(4,
      h4("Column Deletion Log"),
      verbatimTextOutput("col_delete_log"),
      
      br(),
      actionButton("clear_col_delete_log", "Clear Log", class = "btn-warning btn-sm"),
      
      br(), br(),
      h4("Column Summary"),
      verbatimTextOutput("col_delete_summary")
    )
  )
)

server <- function(input, output, session) {
  # Initial data with more columns for deletion testing
  initial_data <- data.frame(
    ID = 1:5,
    Name = paste("Person", 1:5),
    Age = sample(25:45, 5),
    Department = sample(c("IT", "HR", "Finance"), 5, replace = TRUE),
    Salary = sample(40000:80000, 5),
    Bonus = sample(1000:5000, 5),
    Active = sample(c(TRUE, FALSE), 5, replace = TRUE)
  )
  
  # Reactive values
  col_delete_log <- reactiveVal(character(0))
  original_col_names <- names(initial_data)
  
  output$remove_col_table <- renderHandsontable({
    handsontable(initial_data) |>
      hot_col(col = 1, readOnly = TRUE) |>
      hot_col(col = 3, type = "numeric") |>
      hot_col(col = 4, type = "dropdown", source = c("IT", "HR", "Finance")) |>
      hot_col(col = 5, type = "numeric", format = "$0,0") |>
      hot_col(col = 6, type = "numeric", format = "$0,0") |>
      hot_col(col = 7, type = "checkbox") |>
      hot_table(contextMenu = TRUE)
  })
  
  # Observe table changes
  observeEvent(input$remove_col_table, {
    if (!is.null(input$remove_col_table)) {
      # Check if this is an afterRemoveCol event
      if (!is.null(input$remove_col_table$event) && 
          input$remove_col_table$event == "afterRemoveCol") {
        
        # Extract event details
        index <- input$remove_col_table$index + 1  # Convert to 1-based
        amount <- input$remove_col_table$amount
        remaining_colnames <- input$remove_col_table$colnames
        
        # Create log entry
        timestamp <- format(Sys.time(), "%H:%M:%S")
        log_entry <- paste0(
          "[", timestamp, "] EVENT: afterRemoveCol\n",
          "  Removed ", amount, " column(s) starting at position ", index, "\n",
          "  Remaining columns: ", length(remaining_colnames), "\n",
          "  Column names: ", paste(remaining_colnames, collapse = ", ")
        )
        
        # Add to log
        current_log <- col_delete_log()
        new_log <- c(log_entry, current_log)
        if (length(new_log) > 10) {
          new_log <- new_log[1:10]
        }
        col_delete_log(new_log)
      }
    }
  })
  
  # Display deletion log
  output$col_delete_log <- renderText({
    log_entries <- col_delete_log()
    if (length(log_entries) == 0) {
      "No columns deleted yet. Right-click and select 'Delete column'."
    } else {
      paste(log_entries, collapse = "\n\n---\n\n")
    }
  })
  
  # Clear log button
  observeEvent(input$clear_col_delete_log, {
    col_delete_log(character(0))
  })
  
  # Display deletion summary
  output$col_delete_summary <- renderText({
    current_cols <- if (!is.null(input$remove_col_table$colnames)) {
      input$remove_col_table$colnames
    } else {
      original_col_names
    }
    
    deleted_count <- length(original_col_names) - length(current_cols)
    
    paste0(
      "Original columns: ", length(original_col_names), "\n",
      "Current columns: ", length(current_cols), "\n",
      "Deleted columns: ", deleted_count, "\n",
      "Remaining: ", paste(current_cols, collapse = ", ")
    )
  })
}

shinyApp(ui, server)
```

</details>

### Hook Event Summary

All hooks provide consistent event information:

- **`event`**: The hook name (e.g., "afterChange", "afterSelection")
- **`data`**: Current table data
- **`colnames`**: Current column names
- **Hook-specific parameters**: Additional context like `changes`, `source`, `index`, `amount`, etc.

This event system allows you to create sophisticated reactive applications that respond precisely to user interactions with your handsontables.

This covers the essential patterns for using handsontables in Shiny 
applications. The next vignette will cover data validation and advanced 
customization options.

## Interactive Data Editing

### Editable Table with Validation

```{r editable-validation}
library(shiny)
library(handsontable)

ui <- fluidPage(
  titlePanel("Editable Table with Validation"),
  handsontableOutput("table"),
  br(),
  actionButton("save", "Save Changes", class = "btn-primary"),
  actionButton("reset", "Reset Data", class = "btn-warning"),
  br(),
  br(),
  verbatimTextOutput("status")
)

server <- function(input, output, session) {
  # Original data
  original_data <- data.frame(
    ID = 1:5,
    Name = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
    Email = c(
      "alice@company.com",
      "bob@company.com",
      "charlie@company.com",
      "diana@company.com",
      "eve@company.com"
    ),
    Salary = c(50000, 60000, 55000, 65000, 58000),
    Department = c("IT", "HR", "IT", "Finance", "HR")
  )

  # Current working data
  current_data <- reactiveVal(original_data)

  output$table <- renderHandsontable({
    handsontable(current_data()) |>
      hot_col(col = 1, readOnly = TRUE) |> # ID is read-only
      hot_col(col = 2, type = "text") |>
      hot_col(col = 3, type = "text") |>
      hot_validate(
        col = 3,
        type = "regexp",
        pattern = "^[\\w\\._%+-]+@[\\w\\.-]+\\.[A-Za-z]{2,}$"
      ) |>
      hot_col(col = 4, type = "numeric", format = "$0,0") |>
      hot_validate(col = 4, type = "numeric", min = 30000, max = 100000) |>
      hot_col(
        col = 5,
        type = "dropdown",
        source = c("IT", "HR", "Finance", "Marketing"),
        strict = TRUE
      ) |>
      hot_table(contextMenu = TRUE)
  })

  # Update current data when table changes
  observeEvent(input$table, {
    if (!is.null(input$table)) {
      current_data(hot_to_r(input$table))
    }
  })

  # Save changes
  observeEvent(input$save, {
    # In a real app, you would save to database here
    showNotification("Data saved successfully!", type = "success")
  })

  # Reset to original data
  observeEvent(input$reset, {
    current_data(original_data)
    showNotification("Data reset to original values", type = "info")
  })

  output$status <- renderText({
    data_changed <- !identical(current_data(), original_data)
    if (data_changed) {
      "⚠️ Data has been modified"
    } else {
      "✅ Data matches original"
    }
  })
}

shinyApp(ui, server)
```

## Programmatic Updates

### Updating Cell Values

```{r programmatic-updates}
library(shiny)
library(handsontable)

ui <- fluidPage(
  titlePanel("Programmatic Updates"),
  fluidRow(
    column(6, handsontableOutput("table")),
    column(
      6,
      h4("Update Controls"),
      numericInput("row", "Row:", value = 1, min = 1, max = 5),
      numericInput("col", "Column:", value = 1, min = 1, max = 3),
      textInput("value", "New Value:", value = ""),
      actionButton("update", "Update Cell"),
      br(),
      br(),
      actionButton("random", "Fill Random Data"),
      actionButton("clear", "Clear All")
    )
  )
)

server <- function(input, output, session) {
  table_data <- reactiveVal(data.frame(
    A = rep("", 5),
    B = rep("", 5),
    C = rep("", 5)
  ))

  output$table <- renderHandsontable({
    handsontable(table_data())
  })

  # Update specific cell
  observeEvent(input$update, {
    req(input$row, input$col, input$value)
    set_data("table", input$row, input$col, input$value)
  })

  # Fill with random data
  observeEvent(input$random, {
    new_data <- data.frame(
      A = sample(letters[1:10], 5, replace = TRUE),
      B = sample(1:100, 5),
      C = sample(c("X", "Y", "Z"), 5, replace = TRUE)
    )
    table_data(new_data)
  })

  # Clear all data
  observeEvent(input$clear, {
    table_data(data.frame(
      A = rep("", 5),
      B = rep("", 5),
      C = rep("", 5)
    ))
  })

  # Track changes
  observeEvent(input$table, {
    if (!is.null(input$table)) {
      table_data(hot_to_r(input$table))
    }
  })
}

shinyApp(ui, server)
```

## Advanced Shiny Patterns

### Master-Detail Views

```{r master-detail}
library(shiny)
library(handsontable)

ui <- fluidPage(
  titlePanel("Master-Detail Pattern"),
  fluidRow(
    column(8, h4("Employees"), handsontableOutput("employees")),
    column(4, h4("Selected Employee Details"), uiOutput("employee_details"))
  )
)

server <- function(input, output, session) {
  employees <- data.frame(
    ID = 1:10,
    Name = paste("Employee", 1:10),
    Department = sample(c("IT", "HR", "Finance"), 10, replace = TRUE),
    Salary = sample(40000:80000, 10)
  )

  output$employees <- renderHandsontable({
    handsontable(employees) |>
      hot_col(col = 1, readOnly = TRUE) |>
      hot_col(col = 4, type = "numeric", format = "$0,0") |>
      hot_table(contextMenu = TRUE)
  })

  # Get selected employee
  selected_employee <- reactive({
    if (!is.null(input$employees_select)) {
      row_index <- input$employees_select$row + 1 # Convert to 1-based
      if (row_index <= nrow(employees)) {
        employees[row_index, ]
      } else {
        NULL
      }
    } else {
      NULL
    }
  })

  output$employee_details <- renderUI({
    emp <- selected_employee()
    if (is.null(emp)) {
      p("Select an employee from the table")
    } else {
      div(
        h5(paste("Employee ID:", emp$ID)),
        p(strong("Name:"), emp$Name),
        p(strong("Department:"), emp$Department),
        p(strong("Salary:"), scales::dollar(emp$Salary)),
        br(),
        actionButton("edit_emp", "Edit Employee", class = "btn-primary")
      )
    }
  })
}

shinyApp(ui, server)
```

### Dynamic Column Generation

```{r dynamic-columns}
library(shiny)
library(handsontable)

ui <- fluidPage(
  titlePanel("Dynamic Columns"),
  fluidRow(
    column(
      3,
      h4("Controls"),
      numericInput(
        "num_cols",
        "Number of Columns:",
        value = 3,
        min = 1,
        max = 10
      ),
      textInput("col_prefix", "Column Prefix:", value = "Col"),
      actionButton("generate", "Generate Table")
    ),
    column(9, handsontableOutput("dynamic_table"))
  )
)

server <- function(input, output, session) {
  table_data <- reactiveVal(NULL)

  observeEvent(input$generate, {
    num_cols <- input$num_cols
    col_names <- paste0(input$col_prefix, 1:num_cols)

    # Generate random data
    new_data <- data.frame(
      matrix(
        sample(1:100, 50 * num_cols, replace = TRUE),
        nrow = 50,
        ncol = num_cols
      )
    )
    names(new_data) <- col_names

    table_data(new_data)
  })

  output$dynamic_table <- renderHandsontable({
    req(table_data())

    ht <- handsontable(table_data()) |>
      hot_table(
        contextMenu = TRUE,
        filters = TRUE,
        manualColumnSorting = TRUE
      )

    # Configure each column as numeric
    for (i in seq_len(ncol(table_data()))) {
      ht <- ht |> hot_col(col = i, type = "numeric", format = "0,0")
    }

    ht
  })

  # Initialize with default table
  observeEvent(
    TRUE,
    {
      table_data(data.frame(
        Col1 = 1:10,
        Col2 = 11:20,
        Col3 = 21:30
      ))
    },
    once = TRUE
  )
}

shinyApp(ui, server)
```

## Performance Tips for Shiny

### 1. Debouncing Changes
For large tables, consider debouncing change events:

```{r debouncing, eval=FALSE}
# Use a reactive timer to batch updates
values <- reactiveValues(data = NULL, last_update = Sys.time())

observe({
  invalidateLater(500) # Check every 500ms

  if (
    !is.null(input$table) &&
      difftime(Sys.time(), values$last_update, units = "secs") > 0.5
  ) {
    values$data <- hot_to_r(input$table)
    values$last_update <- Sys.time()
  }
})
```

### 2. Selective Updates
Only update what's necessary:

```{r selective-updates, eval=FALSE}
observeEvent(input$table, {
  # Only process if data actually changed
  new_data <- hot_to_r(input$table)
  if (!identical(new_data, previous_data())) {
    # Process changes
    previous_data(new_data)
  }
})
```

### 3. Large Dataset Handling
For very large datasets:

```{r large-datasets, eval=FALSE}
# Use server-side filtering and pagination
# Limit initial display to reasonable size
display_data <- reactive({
  full_data()[1:min(1000, nrow(full_data())), ]
})

output$table <- renderHandsontable({
  handsontable(display_data()) |>
    hot_table(
      filters = FALSE, # Disable client-side filtering
      manualColumnSorting = FALSE # Handle sorting server-side
    )
})
```

## Integration with Other Packages

### With DT Package

```{r dt-integration, eval=FALSE}
# Use handsontable for editing, DT for display
library(shiny)
library(handsontable)
library(DT)

# In UI
fluidPage(
  tabsetPanel(
    tabPanel("Edit", handsontableOutput("edit_table")),
    tabPanel("View", DT::dataTableOutput("view_table"))
  )
)

# In server
shared_data <- reactiveVal(mtcars)

output$edit_table <- renderHandsontable({
  handsontable(shared_data())
})

output$view_table <- DT::renderDataTable({
  DT::datatable(shared_data(), options = list(pageLength = 25))
})

observeEvent(input$edit_table, {
  if (!is.null(input$edit_table)) {
    shared_data(hot_to_r(input$edit_table))
  }
})
```
