---
title: "Data Validation and Cell Types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Validation and Cell Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = interactive()
)
```

```{r setup}
library(handsontable)
library(shiny)
```

# Complete Cell Type Support

Handsontable R now supports all **10 cell types** from Handsontable JS v6.2.2 with full validation and customization options.

## All Available Cell Types

```{r all-cell-types}
# Demonstrate all cell types in one table
demo_data <- data.frame(
  text_col = c("Hello", "World", "Text"),
  numeric_col = c(1234.56, 2345.67, 3456.78),
  date_col = as.Date(c("2024-01-01", "2024-02-01", "2024-03-01")),
  time_col = c("14:30:00", "09:15:30", "18:45:15"),
  dropdown_col = c("Option A", "Option B", "Option C"),
  autocomplete_col = c("Apple", "Banana", "Cherry"),
  checkbox_col = c(TRUE, FALSE, TRUE),
  password_col = c("secret123", "hidden456", "private789"),
  stringsAsFactors = FALSE
)

handsontable(demo_data) |>
  hot_col(col = 1, type = "text") |>
  hot_col(col = 2, type = "numeric", numericFormat = list(pattern = "$0,0.00")) |>
  hot_col(col = 3, type = "date", dateFormat = "YYYY-MM-DD") |>
  hot_col(col = 4, type = "time") |>
  hot_col(col = 5, type = "dropdown", source = c("Option A", "Option B", "Option C", "Option D")) |>
  hot_col(
    col = 6, type = "autocomplete",
    source = c("Apple", "Banana", "Cherry", "Date", "Elderberry"),
    filter = TRUE, trimDropdown = TRUE
  ) |>
  hot_col(col = 7, type = "checkbox") |>
  hot_col(col = 8, type = "password", copyable = FALSE)
```

# Enhanced Numeric Formatting

## Advanced Number Patterns

```{r numeric-formatting}
financial_data <- data.frame(
  Product = c("Widget A", "Widget B", "Widget C"),
  Price = c(1234.56, 9876.54, 5432.10),
  Discount = c(0.15, 0.08, 0.22),
  Quantity = c(1000, 2500, 750),
  Revenue = c(1049376, 9047580, 4237478)
)

handsontable(financial_data) |>
  hot_col(col = 2, type = "numeric", numericFormat = list(pattern = "$0,0.00")) |>
  hot_col(col = 3, type = "numeric", numericFormat = list(pattern = "0.0%")) |>
  hot_col(col = 4, type = "numeric", numericFormat = list(pattern = "0,0")) |>
  hot_col(col = 5, type = "numeric", numericFormat = list(pattern = "$0.0a"))
```

# Smart Autocomplete Features

## Filtered Autocomplete with Custom Options

```{r autocomplete-advanced}
inventory_data <- data.frame(
  Category = c("Electronics", "Clothing", "Books"),
  Product = c("Laptop", "T-Shirt", "Novel"),
  Supplier = c("TechCorp", "FashionInc", "BookHouse")
)

handsontable(inventory_data) |>
  hot_col(
    col = 1, type = "autocomplete",
    source = c("Electronics", "Clothing", "Books", "Home & Garden", "Sports", "Automotive"),
    filter = TRUE,
    trimDropdown = TRUE,
    allowEmpty = FALSE
  ) |>
  hot_col(
    col = 3, type = "autocomplete",
    source = c("TechCorp", "FashionInc", "BookHouse", "GlobalSupply", "LocalVendor", "DirectManufacturer"),
    filter = TRUE,
    strict = FALSE
  )
```

# Date and Time Configuration

## Enhanced Date Picker

```{r date-configuration}
schedule_data <- data.frame(
  Event = c("Meeting", "Presentation", "Workshop"),
  Start_Date = as.Date(c("2024-01-15", "2024-02-20", "2024-03-10")),
  End_Date = as.Date(c("2024-01-15", "2024-02-21", "2024-03-12")),
  Time = c("09:00", "14:30", "10:00")
)

handsontable(schedule_data) |>
  hot_col(
    col = 2, type = "date",
    dateFormat = "YYYY-MM-DD",
    defaultDate = "2024-01-01"
  ) |>
  hot_col(
    col = 3, type = "date",
    dateFormat = "YYYY-MM-DD"
  ) |>
  hot_col(col = 4, type = "time")
```

# Comprehensive Validation System

## All Validation Types

```{r comprehensive-validation}
# Employee data with all validation types
employee_data <- data.frame(
  ID = c("EMP001", "EMP002", "EMP003"),
  Name = c("John Doe", "Jane Smith", "Bob Johnson"),
  Salary = c(75000, 82000, 69000),
  Department = c("Engineering", "Marketing", "Sales"),
  Email = c("john@company.com", "jane@company.com", "bob@company.com"),
  Phone = c("555-123-4567", "555-987-6543", "555-555-0123"),
  Rating = c(4.5, 3.8, 4.2),
  Active = c(TRUE, TRUE, FALSE)
)

handsontable(employee_data) |>
  # Column types
  hot_col(col = 1, type = "text") |>
  hot_col(col = 2, type = "text") |>
  hot_col(col = 3, type = "numeric", numericFormat = list(pattern = "$0,0")) |>
  hot_col(
    col = 4, type = "dropdown",
    source = c("Engineering", "Marketing", "Sales", "HR", "Finance")
  ) |>
  hot_col(col = 5, type = "text") |>
  hot_col(col = 6, type = "text") |>
  hot_col(col = 7, type = "numeric", numericFormat = list(pattern = "0.0")) |>
  hot_col(col = 8, type = "checkbox") |>
  # Validation rules
  hot_validate(cols = 1, type = "regexp", pattern = "^EMP\\d{3}$") |>
  hot_validate(cols = 3, type = "numeric", min = 30000, max = 200000) |>
  hot_validate(
    cols = 4, type = "list",
    source = c("Engineering", "Marketing", "Sales", "HR", "Finance")
  ) |>
  hot_validate(
    cols = 5, type = "regexp",
    pattern = "^[\\w\\._%+-]+@[\\w\\.-]+\\.[A-Za-z]{2,}$"
  ) |>
  hot_validate(cols = 6, type = "regexp", pattern = "^\\d{3}-\\d{3}-\\d{4}$") |>
  hot_validate(cols = 7, type = "numeric", min = 1.0, max = 5.0)
```

# Password and Security

## Password Column with Security Options

```{r password-security}
user_data <- data.frame(
  Username = c("admin", "user1", "user2"),
  Password = c("admin123", "pass456", "secret789"),
  Role = c("Administrator", "User", "Guest"),
  Access_Level = c(3, 2, 1)
)

handsontable(user_data) |>
  hot_col(col = 2, type = "password", copyable = FALSE) |>
  hot_col(
    col = 3, type = "dropdown",
    source = c("Administrator", "User", "Guest", "Moderator")
  ) |>
  hot_col(col = 4, type = "numeric") |>
  hot_validate(cols = 4, type = "numeric", min = 1, max = 5)
```

# Complete Example: Order Management

```{r complete-example}
# Real-world order management example
order_data <- data.frame(
  Order_ID = c("ORD-001", "ORD-002", "ORD-003"),
  Customer = c("TechCorp Inc", "StartupXYZ", "Enterprise Ltd"),
  Product = c("Premium License", "Basic Package", "Enterprise Suite"),
  Quantity = c(50, 100, 25),
  Unit_Price = c(199.99, 99.99, 999.99),
  Discount = c(0.10, 0.05, 0.15),
  Order_Date = as.Date(c("2024-01-15", "2024-01-20", "2024-01-25")),
  Delivery_Date = as.Date(c("2024-01-30", "2024-02-05", "2024-02-10")),
  Status = c("Processing", "Shipped", "Delivered"),
  Priority = c("High", "Medium", "Critical"),
  Notes = c("Rush order", "Standard delivery", "VIP customer"),
  Confirmed = c(TRUE, TRUE, FALSE)
)

handsontable(order_data, height = "400px") |>
  # Column configurations with all features
  hot_col(col = 1, type = "text", readOnly = TRUE, width = 80) |>
  hot_col(
    col = 2,
    type = "autocomplete",
    source = c(
      "TechCorp Inc",
      "StartupXYZ",
      "Enterprise Ltd",
      "NewClient Co",
      "BigCorp"
    ),
    filter = TRUE,
    trimDropdown = TRUE
  ) |>
  hot_col(
    col = 3,
    type = "dropdown",
    source = c(
      "Premium License",
      "Basic Package",
      "Enterprise Suite",
      "Starter Pack"
    )
  ) |>
  hot_col(col = 4, type = "numeric", numericFormat = list(pattern = "0,0")) |>
  hot_col(
    col = 5,
    type = "numeric",
    numericFormat = list(pattern = "$0,0.00")
  ) |>
  hot_col(col = 6, type = "numeric", numericFormat = list(pattern = "0.0%")) |>
  hot_col(col = 7, type = "date", dateFormat = "YYYY-MM-DD") |>
  hot_col(col = 8, type = "date", dateFormat = "YYYY-MM-DD") |>
  hot_col(
    col = 9,
    type = "dropdown",
    source = c("Processing", "Shipped", "Delivered", "Cancelled", "Pending")
  ) |>
  hot_col(
    col = 10,
    type = "dropdown",
    source = c("Low", "Medium", "High", "Critical")
  ) |>
  hot_col(col = 11, type = "text", width = 150) |>
  hot_col(col = 12, type = "checkbox") |>
  # Comprehensive validation
  hot_validate(cols = 1, type = "regexp", pattern = "^ORD-\\d{3}$") |>
  hot_validate(cols = 4, type = "numeric", min = 1, max = 10000) |>
  hot_validate(cols = 5, type = "numeric", min = 0.01, max = 99999.99) |>
  hot_validate(cols = 6, type = "numeric", min = 0, max = 0.50) |>
  hot_validate(
    cols = 9,
    type = "list",
    source = c("Processing", "Shipped", "Delivered", "Cancelled", "Pending")
  ) |>
  hot_validate(
    cols = 10,
    type = "list",
    source = c("Low", "Medium", "High", "Critical")
  ) |>
  # Table features
  handsontable(
    contextMenu = TRUE,
    search = TRUE,
    filters = TRUE,
    manualColumnSorting = TRUE,
    undo = TRUE
  ) |>
  hot_cols(
    manualColumnResize = TRUE,
    fixedColumnsLeft = 1
  )
```

# Shiny Integration Example

```{r shiny-example, eval=FALSE}
# Complete Shiny app with all cell types and validation
ui <- fluidPage(
  titlePanel("Complete Cell Types Demo"),
  fluidRow(
    column(
      12,
      handsontableOutput("demo_table"),
      br(),
      verbatimTextOutput("table_output")
    )
  )
)

server <- function(input, output, session) {
  # Sample data with all types
  demo_data <- data.frame(
    text_col = c("Sample", "Text", "Data"),
    numeric_col = c(1234.56, 2345.67, 3456.78),
    date_col = as.Date(c("2024-01-01", "2024-02-01", "2024-03-01")),
    dropdown_col = c("A", "B", "C"),
    autocomplete_col = c("Apple", "Banana", "Cherry"),
    checkbox_col = c(TRUE, FALSE, TRUE),
    password_col = c("secret1", "secret2", "secret3"),
    stringsAsFactors = FALSE
  )

  output$demo_table <- renderHandsontable({
    handsontable(demo_data) |>
      hot_col(col = 1, type = "text") |>
      hot_col(
        col = 2,
        type = "numeric",
        numericFormat = list(pattern = "$0,0.00")
      ) |>
      hot_col(col = 3, type = "date", dateFormat = "YYYY-MM-DD") |>
      hot_col(col = 4, type = "dropdown", source = c("A", "B", "C", "D")) |>
      hot_col(
        col = 6,
        type = "autocomplete",
        source = c("Apple", "Banana", "Cherry", "Date", "Elderberry"),
        filter = TRUE,
        trimDropdown = TRUE
      ) |>
      hot_col(col = 6, type = "checkbox") |>
      hot_col(col = 7, type = "password", copyable = FALSE) |>
      hot_validate(cols = 2, type = "numeric", min = 0, max = 10000)
  })

  output$table_output <- renderText({
    if (!is.null(input$demo_table)) {
      paste("Table updated at:", Sys.time())
    }
  })
}

shinyApp(ui, server)
```

# Best Practices

## Type Selection Guide

| Data Type | Recommended Cell Type | Key Features |
|-----------|----------------------|--------------|
| Short text | `text` | Basic text input |
| Long text | `text` | Multi-line support |
| Currency | `numeric` + `numericFormat` | Currency formatting |
| Percentages | `numeric` + `numericFormat` | Percentage display |
| Dates | `date` | Date picker |
| Times | `time` | Time picker |
| Fixed options | `dropdown` | Strict validation |
| Searchable options | `autocomplete` | Filtering + suggestions |
| Yes/No | `checkbox` | Boolean values |
| Sensitive data | `password` | Masked input |

## Validation Strategy

1. **Start with type-level validation** using appropriate cell types
2. **Add `hot_validate()`** for business rules and constraints  
3. **Use `allowInvalid = FALSE`** to prevent invalid entries
4. **Combine multiple validation types** for comprehensive data quality

## Performance Tips

- Use `filter = TRUE` on autocomplete for large option lists
- Set `trimDropdown = TRUE` to optimize screen space
- Apply `readOnly = TRUE` to calculated or system fields
- Use `fixedColumnsLeft` for wide tables with key identifier columns

Start with basic types and validation, then enhance with advanced formatting and features as needed!
